Code used to generate paper results : Database comparison and CT characteristics

### 2024-04-25 Entire_PMD_analysis.Rmd
Updated PMD analysis as it will no longer be in paper
Changed everything to adapt to new gene list

@@

```{r CT_genes_prep}
CT_genes <- CT_genes %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000))

@@

distrib_enriched <- unlist(bplapply(1:1000, random_percentage_chrprop, 
                                    data = enriched_genes_analysable, 
                                    nchrX = PMD_chr_repartition[[2]], 
                                    nautosomes = PMD_chr_repartition[[1]]))
hist(distrib_enriched,
     col="violetred1",
     border = "violetred3",
     breaks = 20,
     main = paste0("Percentage of promoters in PMDs when randomly taking ",
                   sum(table(CT_meth_regulated$PMD)), 
                   " tissue-specific 
genes 1000 times (taking chromosome proportion into account)"),
     cex.main = 0.75,
     xlab = "Percentages",
     ylab = "Frequency",
     xlim = c(20,60))
abline(v = PMD_perc, col = "slateblue1", lwd = 3)
text(PMD_perc, 120, "CT genes regulated 
by methylation", col = "slateblue4", cex = 0.8)
abline(v = CT_not_meth_reg_PMD_perc, col = "darkseagreen1", lwd = 3)
text(CT_not_meth_reg_PMD_perc, 120, "CT genes not regulated 
by methylation", col = "darkseagreen", cex = 0.8)

@@

# nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
#                                       reorganize_data_ensembl, data = nonCTmeth_inCD4)
# names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)
# 
# p <- rep(NA, length(nonCT_per_gene_CD4_datasets))
# for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
#  p[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) > 1 &
#    is.data.frame(nonCT_per_gene_CD4_datasets[[i]])
# }
# nonCT_per_gene_CD4_datasets <- nonCT_per_gene_CD4_datasets[p]
# 
# k <- rep(NA, length(nonCT_per_gene_CD4_datasets))
# for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
#   k[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) == 2
# }
# 
# for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
#   tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
#   nonCT_per_gene_CD4_datasets[[gene]] <- rbind(nonCT_per_gene_CD4_datasets[[gene]], tmp)
#   rownames(nonCT_per_gene_CD4_datasets[[gene]])[3] <-
#     c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in%
#                                                 rownames(nonCT_per_gene_CD4_datasets[[gene]]))]
#   # naming the new row
# }
# 
# for (gene in names(nonCT_per_gene_CD4_datasets)) {
#   nonCT_per_gene_CD4_datasets[[gene]] <- arrange(nonCT_per_gene_CD4_datasets[[gene]],
#                                               rownames(nonCT_per_gene_CD4_datasets[[gene]]))
#   # Put rows in the right order
# }
# 
# rm(p, i, k, tmp)

@@

nonCT_meth_means <- sapply(nonCT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
nonCT_meth_means <- nonCT_meth_means %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id")

>>>>

This should be completed using another model of senescene/immortalization by 
confirming the difference between CTs and PMDs, and CTs and nonCT in PMDs.

@@

```{r CT_genes_prep}
load("../CT_genes_latest_version.rda")
CT_genes <- CT_genes %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000))

@@

distrib_enriched <- unlist(bplapply(1:1000, random_percentage_chrprop, 
                                    data = enriched_genes_analysable, 
                                    nchrX = PMD_chr_repartition[[2]], 
                                    nautosomes = PMD_chr_repartition[[1]]))
hist(distrib_enriched,
     col="violetred1",
     border = "violetred3",
     breaks = 20,
     main = paste0("Percentage of promoters in PMDs when randomly taking",
                   sum(table(CT_meth_regulated$PMD)), 
                   " tissue-specific 
genes 1000 times (taking chromosome proportion into account)"),
     cex.main = 0.75,
     xlab = "Percentages",
     ylab = "Frequency",
     xlim = c(20,70))
abline(v = PMD_perc, col = "slateblue1", lwd = 3)
text(PMD_perc, 75, "CT genes regulated 
by methylation", col = "slateblue4", cex = 0.8)
abline(v = CT_not_meth_reg_PMD_perc, col = "darkseagreen1", lwd = 3)
text(CT_not_meth_reg_PMD_perc, 75, "CT genes not regulated 
by methylation", col = "darkseagreen", cex = 0.8)

@@

# nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
#                                       reorganize_data_ensembl, data = nonCTmeth_inCD4,
#                                       BPPARAM = MulticoreParam(8L))
# names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)
# 
# p <- rep(NA, length(nonCT_per_gene_CD4_datasets))
# for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
#  p[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) > 1 &
#    is.data.frame(nonCT_per_gene_CD4_datasets[[i]])
# }
# nonCT_per_gene_CD4_datasets <- nonCT_per_gene_CD4_datasets[p]
# 
# k <- rep(NA, length(nonCT_per_gene_CD4_datasets))
# for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
#   k[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) == 2
# }
# 
# for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
#   print(gene)
#   tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
#   nonCT_per_gene_CD4_datasets[[gene]] <- rbind(nonCT_per_gene_CD4_datasets[[gene]], tmp)
#   rownames(nonCT_per_gene_CD4_datasets[[gene]])[3] <-
#     c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in%
#                                                 rownames(nonCT_per_gene_CD4_datasets[[gene]]))]
#   # naming the new row
# }
# 
# for (gene in names(nonCT_per_gene_CD4_datasets)) {
#   nonCT_per_gene_CD4_datasets[[gene]] <- arrange(nonCT_per_gene_CD4_datasets[[gene]],
#                                               rownames(nonCT_per_gene_CD4_datasets[[gene]]))
#   # Put rows in the right order
# }
# 
# rm(p, i, k, tmp)
# save(nonCT_per_gene_CD4_datasets, 
#      file = "data_rmd/nonCT_per_gene_CD4_datasets.rda")

@@

nonCT_meth_means_tmp <- sapply(nonCT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
nonCT_meth_means <- Reduce(rbind, nonCT_meth_means_tmp) %>% 
  data.frame()
nonCT_meth_means$ensembl_gene_id <- names(nonCT_meth_means_tmp)

### 2024-04-25 Paper_code-Database_CT.Rmd
Changed everything to adapt to new gene list

```{r load_data, warning=FALSE, message=FALSE}
GTEX_data <- CTdata::GTEX_data()
normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()
CCLE_data <- CTdata::CCLE_data()
TCGA_TPM <- CTdata::TCGA_TPM()
DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
testis_sce <- CTdata::testis_sce()
scRNAseq_HPA <- CTdata::scRNAseq_HPA()
CT_mean_methylation_in_tissues <- CTdata::CT_mean_methylation_in_tissues()
```

@@

lost <- Bruggeman_data[which(!Bruggeman_data$Gene %in% c(Bruggeman_CT$Gene_Name)), "Gene"]
colnames(lost) <- "external_gene_name"
lost$Gene_Name <- rep(NA, 15)
lost$external_synonym <- rep(NA, 15)

lost[lost$external_gene_name == "C21orf59", "Gene_Name"] <- "CFAP298"
lost[lost$external_gene_name == "C11orf57", "Gene_Name"] <- "NKAPD1"
lost[lost$external_gene_name == "C7orf55", "Gene_Name"] <- "FMC1"
lost[lost$external_gene_name == "C10orf12", "Gene_Name"] <- "LCOR"
lost[lost$external_gene_name == "RPL19P12", "Gene_Name"] <- "RPL19P12"
lost[lost$external_gene_name == "C16orf59", "Gene_Name"] <- "TEDC2"
lost[lost$external_gene_name == "TTTY15", "Gene_Name"] <- "USP9Y"
lost[lost$external_gene_name == "C17orf53", "Gene_Name"] <- "HROB"
lost[lost$external_gene_name == "C12orf66", "Gene_Name"] <- "KICS2"
lost[lost$external_gene_name == "C9orf84", "Gene_Name"] <- "SHOC1"
lost[lost$external_gene_name == "C10orf25", "Gene_Name"] <- "ZNF22-AS1"
lost[lost$external_gene_name == "C20orf197", "Gene_Name"] <- "LINC02910"
lost[lost$external_gene_name == "C3orf67", "Gene_Name"] <- "CFAP20DC"
lost[lost$external_gene_name == "C8orf37", "Gene_Name"] <- "CFAP418"
lost[lost$external_gene_name == "C22orf34", "Gene_Name"] <- "MIR3667HG"

@@

missing_Bruggeman <- c("BMS1P4", "ADAM6", "ANXA2P3", "ARHGAP11B", "DPY19L2P2", 
                       "HLA-L", "PA2G4P4", "PIPSL", "PRKY", "YBX3P1", 
                       "RPL23AP53", "UQCRBP1", "RPL23P8", "MRS2P2", "PIN4P1", 
                       "SLC6A10P", "GUSBP2", "PPIEL", "LRRC37BP1", "MSL3P1", 
                       "PLEKHA8P1", "STAG3L1", "TCAM1P", "ZNF702P", "ZNF815P", 
                       "ATP6AP1L", "RPL21P44", "SEC14L1P1", "ZNF876P", 
                       "RPLP0P2", "FAM86JP", "FAM175A", "LACE1", "ATP5EP2", 
                       "WDR92", "TCTE3", "METTL20", "KIAA2022", "ZNRD1-AS1", 
                       "SGOL1", "FAM35DP", "MTL5", "TMEM14E", "MLLT4-AS1", 
                       "CCDC173", "KIAA1524", "WDR78", "LINC00476", "LYRM5", 
                       "HILS1", "CASC5", "KIAA1919", "CTAGE5", "FAM188B", 
                       "TMEM194B", "FAM122C", "PPP1R2P3", "KIAA0391", "SGOL2", 
                       "FAM19A3", "ZNF788", "RPL19P12")

@@

@@

# 10 Genes are lost because not in any database

@@

table(CTdatabase_lost$CCLE_category)
```

## CTexploreR VS omics databases
```{r upset_plot}
core_ours <- Venn(list(Jamin = Jamin_core_CT$Gene, 
                       CTexploreR = CT_genes$external_gene_name))

@@

@@

@@

@@

@@

@@

Most genes are testis specific 
(`r table(CT_genes$testis_specificity)[[2]]/dim(CT_genes)[1]*100`%).
Most genes are mainly protein coding genes 
(`r table(CT_genes$transcript_biotype)[[3]]/dim(CT_genes)[1]*100`%).

@@

On the chromosome X, there is a clear enrichment of CT genes regulated by 
methylation (83/93 chrX genes or 83/160 genes regulated by methylation).

@@

CT_genes %>% 
  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
                                           "Regulated by methylation",
                                           "Not regulated by methylation")) %>% 
  ggplot(aes(x = chr_factor, fill = testis_specificity, color = X_linked)) +
  geom_bar(stat = 'count') +
  scale_fill_manual(values = c("goldenrod1", "mediumpurple1")) +
  scale_color_manual(values = c("royalblue1", "deeppink"), guide = "none") +
  facet_wrap(~ regulated_by_methylation, ncol = 2) +
  theme_bw() +
  xlab("Chromosome") +
  ylab("Number of genes") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "top",
        axis.title = element_text(size = 10, color = "gray10"))
```

@@

### Coactivation in tumours

We know that CT genes that are regulated by methylation are activated in 
tumours due to methylation. However, those that are not it is more blurry. It 
is interesting to look at their expression to see if there are groups of genes 
that are activated together in tumours which would suggest another activation :
a transcription factor ? a gene expresion program from meiosis ? timing of 
expression ? function ?

```{r not_reg}
not_regulated <- filter(CT_genes, !regulated_by_methylation)$external_gene_name
```

As they can blur the clustering, I sometimes remove genes that are in the 
background and  those that are expressed a lot.

- BRCA :

```{r BRCA}
TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "BRCA", units = "log_TPM")
# Removing genes to see clearer

BRCA_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_BRCA > 0.01 & percent_pos_BRCA < 30) %>% 
  pull(external_gene_name)

BRCA_genes_ordered <- 
  BRCA_genes[row_order(draw(TCGA_expression(BRCA_genes, tumor = "BRCA", 
                                            units = "log_TPM")))]

BRCA_cluster1 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "AKAP14"):
                       which(BRCA_genes_ordered == "C11orf97"))]
BRCA_cluster2 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "GTF2I-AS1"):
                       which(BRCA_genes_ordered == "SLIT3-AS1"))]

We can see two small clusters.

- LUAD :

```{r LUAD}
TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "LUAD", units = "log_TPM")
# Removing genes to see clearer
LUAD_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_LUAD > 0.01 & percent_pos_LUAD < 30) %>% 
  pull(external_gene_name)

LUAD_genes_ordered <- 
  LUAD_genes[row_order(draw(TCGA_expression(LUAD_genes, tumor = "LUAD", 
                                            units = "log_TPM")))]

LUAD_cluster1 <- c("TSPY1", "TSPY2")
LUAD_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
LUAD_cluster3 <- LUAD_genes_ordered[c(1:22)]
LUAD_cluster4 <- LUAD_genes_ordered[c(which(LUAD_genes_ordered == "GTF2I-AS1"):
                       which(LUAD_genes_ordered == "SLIT3-AS1"))]
```

22 first genes are clearly quite clustering, there is also a small SPANX cluster,
a small TSPY cluster ans a bigger one around the middle of the graph.

- LUSC :

```{r LUSC}
TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "LUSC", units = "log_TPM")
# Removing genes to see clearer

LUSC_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_LUSC > 0.01 & percent_pos_LUSC < 30) %>% 
  pull(external_gene_name)

LUSC_genes_ordered <- 
  LUSC_genes[row_order(draw(TCGA_expression(LUSC_genes, tumor = "LUSC", 
                                            units = "log_TPM")))]

LUSC_cluster1 <- c("TSPY1", "TSPY2",  "TSPY9", "TSPY10")
LUSC_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
LUSC_cluster3 <- LUSC_genes_ordered[c(1:15)]

We can see a cluster at the top and two smaller ones corresponding to SPANX and
TSPY genes. As well as quite a big cluster with the 15 first lines

- SKCM :

```{r SKCM}
SKCM_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_SKCM > 0.01 & percent_pos_SKCM < 30) %>% 
  pull(external_gene_name)

TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "SKCM", units = "log_TPM")
# Lots of genes activated almost everywhere or nowhere that can mask clusters
TCGA_expression(SKCM_genes,
                tumor = "SKCM", units = "log_TPM")

SKCM_cluster <- c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10")
```

It is clear that there is a TSPY cluster again.

- Comparison

TSPY and SPANX genes are found reactivated/coactivated in different types of 
tumours. Let's compare the bigger groups found in LUAD, LUSC and BRCA

```{r upset_clusters}
upset(fromList(list(LUSC = LUSC_cluster3,
               LUAD1 = LUAD_cluster3,
               LUAD2 = LUAD_cluster4,
               BRCA1 = BRCA_cluster1,
               BRCA2 = BRCA_cluster2)))
```

There is quite a lot of similarities here between the clusters. Let's look at
the expression in tumours.

```{r TCGA_clusters}
TCGA_expression(c(LUSC_cluster3, LUAD_cluster3, LUAD_cluster4, BRCA_cluster1,
                  BRCA_cluster2),
                tumor = "all", units = "log_TPM")
```

Let's have a look at the timings of expression.

```{r timing_cluster, message = FALSE}
testis_expression(c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10"), 
                  cells = "all")
testis_expression(c("SPANXC", "SPANXB1",  "SPANXD"), 
                  cells = "all")
testis_expression(LUSC_cluster3, cells = "all")
testis_expression(LUAD_cluster3, cells = "all")
testis_expression(LUAD_cluster4, cells = "all")
testis_expression(BRCA_cluster1, cells = "all")
testis_expression(BRCA_cluster2, cells = "all")
```

They mainly seem to, in a cluster, be expressed at the same time.

@@

There is only data for 251 of our 298 CT genes.

@@

41% of genes are mainly expressed pre-meioticly.

@@

14 genes are on the X chromosome but escape X chromosome inactivation and are
activates post-meioticly.

@@

@@

# Combine the two cilium movement from BP
ora_to_plot[ora_to_plot$ID == "GO:0003341", "Count"] <- 
  ora_to_plot[ora_to_plot$ID == "GO:0003341", "Count"] +
  ora_to_plot[ora_to_plot$ID == "GO:0060294", "Count"]

ora_to_plot <- ora_to_plot %>% 
  arrange(desc(Count)) %>% 
  head(11) %>% 
  filter(ID != "GO:0060294") %>% 
  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/177,
                           ONTOLOGY == "CC"~ Count/197,
                           ONTOLOGY == "MF"~ Count/186))

@@

As we can see here, most of genes are indeed linked to functions from reproduction.
I represented here the 12 categories with the most genes, all from biological 
processes. However, they are enriched in 89 different GO terms.

@@

```{r GO_ORA_meth}
go_ora_meth <- enrichGO(gene = 
                          filter(CT_genes, regulated_by_methylation)$ensembl_gene_id,
                        keyType = "ENSEMBL",
                        OrgDb = org.Hs.eg.db,
                        ont = "all",
                        readable = TRUE)
go_ora_meth <- simplify(go_ora_meth)

go_ora_not_meth <- enrichGO(gene = 
                          filter(CT_genes, !regulated_by_methylation)$ensembl_gene_id,
                        keyType = "ENSEMBL",
                        OrgDb = org.Hs.eg.db,
                        ont = "all",
                        readable = TRUE)

go_ora_not_meth <- simplify(go_ora_not_meth)
as_tibble(go_ora_not_meth)

go_ora_meth_plot <- as_tibble(go_ora_meth) %>% 
  arrange(desc(Count)) %>% 
  head(11) %>% 
  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/101,
                           ONTOLOGY == "MF"~ Count/107)) %>%  
  mutate(regulation = "Methylation")

go_ora_not_meth_plot <- as_tibble(go_ora_not_meth) %>% 
  arrange(desc(Count)) %>% 
  head(11) %>% 
  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/76,
                           ONTOLOGY == "CC"~ Count/93))%>%  
  mutate(regulation = "Not methylation")

>>>>

```{r load_data, warning=FALSE, message=FALSE}
GTEX_data <- CTdata::GTEX_data()
normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()
CCLE_data <- CTdata::CCLE_data()
TCGA_TPM <- CTdata::TCGA_TPM()
DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
testis_sce <- CTdata::testis_sce()
scRNAseq_HPA <- CTdata::scRNAseq_HPA()
CT_mean_methylation_in_tissues <- CTdata::CT_mean_methylation_in_tissues()
####
load("../CT_genes_latest_version.rda")
CT_genes$testis_specificity <- "testis_specific"
```

@@

lost <- Bruggeman_data[which(!Bruggeman_data$Gene %in% c(Bruggeman_CT$Gene_Name)), "Gene"]
colnames(lost) <- "external_gene_name"
lost$Gene_Name <- rep(NA, nrow(lost))
lost$external_synonym <- rep(NA, nrow(lost))

lost[lost$external_gene_name == "C21orf59", "Gene_Name"] <- "CFAP298"
lost[lost$external_gene_name == "C11orf57", "Gene_Name"] <- "NKAPD1"
lost[lost$external_gene_name == "C7orf55", "Gene_Name"] <- "FMC1"
lost[lost$external_gene_name == "C10orf12", "Gene_Name"] <- "LCOR"
lost[lost$external_gene_name == "RPL19P12", "Gene_Name"] <- "RPL19P12"
lost[lost$external_gene_name == "C16orf59", "Gene_Name"] <- "TEDC2"
lost[lost$external_gene_name == "TTTY15", "Gene_Name"] <- "USP9Y"
lost[lost$external_gene_name == "C17orf53", "Gene_Name"] <- "HROB"
lost[lost$external_gene_name == "C1orf112", "Gene_Name"] <- "FIRRM"
lost[lost$external_gene_name == "C12orf66", "Gene_Name"] <- "KICS2"
lost[lost$external_gene_name == "C9orf84", "Gene_Name"] <- "SHOC1"
lost[lost$external_gene_name == "C10orf25", "Gene_Name"] <- "ZNF22-AS1"
lost[lost$external_gene_name == "C20orf197", "Gene_Name"] <- "LINC02910"
lost[lost$external_gene_name == "C3orf67", "Gene_Name"] <- "CFAP20DC"
lost[lost$external_gene_name == "C8orf37", "Gene_Name"] <- "CFAP418"
lost[lost$external_gene_name == "C22orf34", "Gene_Name"] <- "MIR3667HG"

@@

missing_Bruggeman <- c("BMS1P4", "ADAM6", "ANXA2P3", "ARHGAP11B", "DPY19L2P2", 
                       "HLA-L", "PA2G4P4", "PIPSL", "PRKY", "YBX3P1", 
                       "RPL23AP53", "UQCRBP1", "RPL23P8", "MRS2P2", "PIN4P1", 
                       "SLC6A10P", "GUSBP2", "PPIEL", "LRRC37BP1", "MSL3P1", 
                       "PLEKHA8P1", "STAG3L1", "TCAM1P", "ZNF702P", "ZNF815P", 
                       "ATP6AP1L", "RPL21P44", "SEC14L1P1", "ZNF876P", 
                       "RPLP0P2", "FAM86JP", "FAM175A", "LACE1", "ATP5EP2", 
                       "WDR92", "TCTE3", "METTL20", "KIAA2022", "ZNRD1-AS1", 
                       "SGOL1", "FAM35DP", "MTL5", "TMEM14E", "MLLT4-AS1", 
                       "CCDC173", "KIAA1524", "WDR78", "LINC00476", "LYRM5", 
                       "HILS1", "CASC5", "KIAA1919", "CTAGE5", "FAM188B", 
                       "TMEM194B", "FAM122C", "PPP1R2P3", "KIAA0391", "SGOL2", 
                       "FAM19A3", "ZNF788", "RPL19P12", "FIRRM")

@@

We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
% of CTdatabase in CTexploreR, which is `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
% of our database.

@@

# 12 Genes are lost because not in any database

@@

table(CTdatabase_lost$CCLE_category)

table(CTdatabase_lost$TCGA_category, CTdatabase_lost$CCLE_category)

`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`% of 
these genes are not testis specific.

`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$TCGA_category, CTdatabase_lost$CCLE_category)["activated", "activated"])/nrow(CTdatabase_lost)*100`
% are not properly activated in tumors and/or cancer cell lines.

## CTexploreR VS omics databases

```{r upset_plot}
core_ours <- Venn(list(Jamin = Jamin_core_CT$Gene, 
                       CTexploreR = CT_genes$external_gene_name))

@@

4 in all, 59 in at least 3 databases

@@

We find `r length(core_ours@IntersectionSets[["11"]])/nrow(Jamin_core_CT)*100`
% of CTdatabase in CTexploreR, which is `r length(core_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
% of our database.

`r (nrow(Jamin_lost) - table(Jamin_lost$testis_specificity)[["testis_specific"]])/nrow(Jamin_lost)*100`% of 
these genes are not testis specific.

`r (nrow(Jamin_lost) - table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)["activated", "activated"])/nrow(Jamin_lost)*100`
% are not properly activated in tumors and/or cancer cell lines.

@@

We find `r length(Wang_ours@IntersectionSets[["11"]])/nrow(Wang_CT)*100`
% of CTdatabase in CTexploreR, which is `r length(Wang_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
% of our database.

`r (nrow(Wang_lost) - table(Wang_lost$testis_specificity)[["testis_specific"]])/nrow(Wang_lost)*100`% of 
these genes are not testis specific.

`r (nrow(Wang_lost) - table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)["activated", "activated"])/nrow(Wang_lost)*100`
% are not properly activated in tumors and/or cancer cell lines.

@@

We find `r length(Carter_ours@IntersectionSets[["11"]])/nrow(Carter_CT)*100`
% of CTdatabase in CTexploreR, which is `r length(Carter_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
% of our database.

`r (nrow(Carter_lost) - table(Carter_lost$testis_specificity)[["testis_specific"]])/nrow(Carter_lost)*100`% of 
these genes are not testis specific.

`r (nrow(Carter_lost) - table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)["activated", "activated"])/nrow(Carter_lost)*100`
% are not properly activated in tumors and/or cancer cell lines.

@@

We find `r length(Bruggeman_ours@IntersectionSets[["11"]])/nrow(Bruggeman_CT)*100`
% of CTdatabase in CTexploreR, which is `r length(Bruggeman_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
% of our database.

`r (nrow(Bruggeman_lost) - table(Bruggeman_lost$testis_specificity)[["testis_specific"]])/nrow(Bruggeman_lost)*100`% of 
these genes are not testis specific.

`r (nrow(Bruggeman_lost) - table(Bruggeman_lost$TCGA_category, Bruggeman_lost$CCLE_category)["activated", "activated"])/nrow(Bruggeman_lost)*100`
% are not properly activated in tumors and/or cancer cell lines.

@@

Most genes are testis specific 
(`r table(CT_genes$testis_specificity)[["testis_specific"]]/dim(CT_genes)[1]*100`%).
Most genes are mainly protein coding genes 
(`r table(CT_genes$transcript_biotype)["protein_coding"]/dim(CT_genes)[1]*100`%).

@@

On the chromosome X, there is a clear enrichment of CT genes regulated by 
methylation (`r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$X_linked)[[2]]` 
chrX genes or `r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$regulated_by_methylation)[[2]]` 
genes regulated by methylation).

@@

CT_genes %>% 
  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
                                           "Regulated by methylation",
                                           "Not regulated by methylation")) %>% 
  ggplot(aes(x = chr_factor, fill = X_linked)) +
  geom_bar(stat = 'count') +
  scale_fill_manual(values = c("royalblue1", "deeppink")) +
  facet_wrap(~ regulated_by_methylation, ncol = 2) +
  theme_bw() +
  xlab("Chromosome") +
  ylab("Number of genes") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none",
        axis.title = element_text(size = 10, color = "gray10"))
```

@@

@@

There is only data for `r sum(CT_genes$external_gene_name %in% unique(rownames(testis_sce)))` of our `r nrow(CT_genes)` CT genes.

@@

`r sum(table(CT_genes$testis_cell_type)[c("Early_spermatocyte", "Spermatogonia", "SSC")])/length(genes_avail)*100`
% of genes are mainly expressed pre-meioticly.

@@

@@

These genes are on the X chromosome but escape X chromosome inactivation and are
activates post-meioticly.

@@

ora_to_plot <- ora_to_plot %>% 
  arrange(desc(Count)) %>% 
  head(9) %>% 
  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/102,
                           ONTOLOGY == "MF"~ Count/186))

@@

As we can see here, most of genes are indeed linked to functions from reproduction.
I represented here the 12 categories with the most genes, all from biological 
processes. However, they are enriched in `r nrow(as_tibble(go_ora))` different GO terms.

@@

```{r GO_ORA_meth}
go_ora_meth <- enrichGO(gene = 
                          filter(CT_genes, regulated_by_methylation)$ensembl_gene_id,
                        keyType = "ENSEMBL",
                        OrgDb = org.Hs.eg.db,
                        ont = "all",
                        readable = TRUE)
go_ora_meth <- simplify(go_ora_meth)
as_tibble(go_ora_meth)

go_ora_not_meth <- enrichGO(gene = 
                          filter(CT_genes, !regulated_by_methylation)$ensembl_gene_id,
                        keyType = "ENSEMBL",
                        OrgDb = org.Hs.eg.db,
                        ont = "all",
                        readable = TRUE)
go_ora_not_meth <- simplify(go_ora_not_meth)
as_tibble(go_ora_not_meth)

go_ora_meth_plot <- as_tibble(go_ora_meth) %>% 
  arrange(desc(Count)) %>% 
  head(10) %>% 
  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/73,
                           ONTOLOGY == "MF"~ Count/80,
                           ONTOLOGY == "CC"~ Count/78)) %>%  
  mutate(regulation = "Methylation")

go_ora_not_meth_plot <- as_tibble(go_ora_not_meth) %>% 
  arrange(desc(Count)) %>% 
  head(8) %>% 
  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/29,
                           ONTOLOGY == "MF"~ Count/30))%>%  
  mutate(regulation = "Not methylation")
